---
title: "R0 Estimation"
output: html_notebook
---

```{r libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(lubridate)
library(knitr)
library(arsenal)
library(R0)
```

Here I take the [R0 package](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3582628/) to estimate the reproduction number for COVID-19.

Load the data from the [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by Johns Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19):

```{r load-data}
confirmed <- read.csv(url("https://raw.githubusercontent.com/CSSEGISandData/2019-nCoV/master/time_series/time_series_2019-ncov-Confirmed.csv"))
```

This data contains a row per state/province and a column per reported case numbers - on a particular date and time. There may be more than report per day at different times for a given location. 

Let's transform this into a daily count using the last total count for each day:

```{r tidy-data}
sums <- colSums(confirmed[,-match(c("Province.State", "Country.Region", "Lat", "Long"), names(confirmed))], na.rm=TRUE)

sumsByDateCode <- as.data.frame(t(t(sums)))
colnames(sumsByDateCode) <- c("count")

sumsByDateCode$datecode <- rownames(sumsByDateCode)

sumsByDate <- mutate(sumsByDateCode, datetime = mdy_hm(substring(datecode,2)))

dailyCounts <- aggregate(sumsByDate$count, by=list(as.Date(sumsByDate$datetime)), FUN=tail, n=1)

colnames(dailyCounts) <- c("date", "count")

dailyCounts$day <- seq.int(nrow(dailyCounts))
```


Now use the R0 package to estimate the reproduction number. I need a generation time, which we don't have to hand, so I take the values use for SARS - a mean of 8.4, standard deviation 3.8 - as [described here](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3816335/).

```{r estimates, warning=FALSE, message=FALSE}
mgt <- generation.time("gamma", c(8.4, 3.8))

est <- estimate.R(dailyCounts$count, methods=c("TD", "EG", "ML", "SB"), GT=mgt)
```

Here is the range of the reproduction number thus estimated using the "time dependendent" method:

```{r merge-data, warning=FALSE, message=FALSE, echo=FALSE}
dailyCountAndPrediction <- merge(dailyCounts, est$estimates$TD$pred, by="row.names", sort=FALSE, all=TRUE)
names(dailyCountAndPrediction)[5] <- "prediction"

estimatedR <- est$estimates$TD$R[1:length(est$estimates$TD$R)-1]
summary(estimatedR)
```

We can plot the actual and predicted values using the "time dependent" method:

```{r plot-predictions, echo=FALSE}
plot(dailyCountAndPrediction$day, dailyCountAndPrediction$count, xlab="days",ylab="count")
lines(dailyCountAndPrediction$day, dailyCountAndPrediction$prediction, col="green")
```

Finally, here is a plot of estimated reproduction number (using the time-dependent method) over time:

```{r plot-estimates, echo=FALSE}
plot(estimatedR, xlab="days", ylab="R", ylim=c(0,20), yaxt="n")
abline(h=1, col="gray60")
axis(2, at=seq(0:max(estimatedR)))
```

