---
title: "R0 Estimation"
author: "Nick Cotter"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file=file.path(dirname(inputFile), 'docs', 'index.html')) })
output: 
      html_document:
        keep_md: yes
---

```{r libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(lubridate)
library(dplyr)
library(knitr)
library(arsenal)
library(R0)
```

Here I take the [R0 package](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3582628/) to estimate the reproduction number for COVID-19. NOTE: my aim is to investigate the R0 package rather than provide sound predictions of COVID-19's future behaviour. In particular I have guessed the generation time.


Load the data from the [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by Johns Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19):



```{r load-data}
confirmed <- read.csv(url("https://raw.githubusercontent.com/CSSEGISandData/2019-nCoV/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv"))
```

This data contains a row per state/province and a column per reported case numbers - on a particular date. Let's tidy this into count per day number since the start.

```{r tidy-data}

confirmed <- subset(confirmed, select = -c(Lat, Long))

sums <- colSums(confirmed[,-match(c("Province.State", "Country.Region"), names(confirmed))], na.rm=TRUE)

sumsByDateCode <- as.data.frame(t(t(sums)))
colnames(sumsByDateCode) <- c("count")

sumsByDateCode$datecode <- rownames(sumsByDateCode)

dailyCounts <- mutate(sumsByDateCode, date = mdy(substring(datecode,2)))

dailyCounts$day <- seq.int(nrow(dailyCounts))
```


Now use the R0 package to estimate the reproduction number. I need a generation time, which we don't have to hand, so I take the values use for SARS - a mean of 8.4, standard deviation 3.8 - as [described here](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3816335/). The different estimation methods (time dependent, exponential growth etc.) are described there too.

```{r estimates, warning=FALSE, message=FALSE}
mgt <- generation.time("gamma", c(8.4, 3.8))

est <- estimate.R(dailyCounts$count, methods=c("TD", "EG", "ML", "SB"), GT=mgt)
```


```{r merge-data, warning=FALSE, message=FALSE, echo=FALSE}
dailyCountAndPrediction <- merge(dailyCounts, est$estimates$TD$pred, by="row.names", sort=FALSE, all=TRUE)
dailyCountAndPrediction <- dailyCountAndPrediction[-c(1)]
names(dailyCountAndPrediction)[5] <- "TD"

dailyCountAndPrediction <- merge(dailyCountAndPrediction, est$estimates$EG$pred, by="row.names", sort=FALSE, all=TRUE)
dailyCountAndPrediction <- dailyCountAndPrediction[-c(1)]
names(dailyCountAndPrediction)[6] <- "EG"

dailyCountAndPrediction <- merge(dailyCountAndPrediction, est$estimates$ML$pred, by="row.names", sort=FALSE, all=TRUE)
dailyCountAndPrediction <- dailyCountAndPrediction[-c(1)]
names(dailyCountAndPrediction)[7] <- "ML"

dailyCountAndPrediction <- merge(dailyCountAndPrediction, est$estimates$SB$pred, by="row.names", sort=FALSE, all=TRUE)
dailyCountAndPrediction <- dailyCountAndPrediction[-c(1)]
names(dailyCountAndPrediction)[8] <- "SB"
```

```{r add-rmse, echo=FALSE}

RMSE = function(m, o){
  sqrt(mean((m - o)^2))
}

methodError <- dailyCountAndPrediction %>%
                  summarise(TD=RMSE(TD,count), EG=RMSE(EG,count), ML=RMSE(ML,count), SB=RMSE(SB,count))

```



We can plot the actual and predicted values using the different estimation methods:

```{r plot-predictions, echo=FALSE}
plot(dailyCountAndPrediction$day, dailyCountAndPrediction$count, xlab="days",ylab="count", col="red")
lines(dailyCountAndPrediction$day, dailyCountAndPrediction$TD, col="green")
lines(dailyCountAndPrediction$day, dailyCountAndPrediction$EG, col="blue")
lines(dailyCountAndPrediction$day, dailyCountAndPrediction$ML, col="purple")
lines(dailyCountAndPrediction$day, dailyCountAndPrediction$SB, col="orange")
legend(1, max(dailyCountAndPrediction$count)-10, legend=c("Actual", "TD", "EG", "ML", "SB"), col=c("red", "green", "blue", "purple", "orange"), lty=c(0,1,1,1,1), pch=c(1,NA,NA,NA,NA), cex=0.8)
```

The time-dependent method seems to fit the best. Here are the RMSE values for the different methods:

```{r summarise-error, echo=FALSE}

kable(methodError)
```


Here is the range of the reproduction number thus estimated using the "time dependendent" method:

```{r td-estimate-summary, warning=FALSE, message=FALSE, echo=FALSE}
estimatedR <- est$estimates$TD$R[1:length(est$estimates$TD$R)-1]
summary(estimatedR)
```

Finally, here is a plot of estimated reproduction number (using the time-dependent method) over time:

```{r plot-estimates, echo=FALSE}
estDf <- as.data.frame(estimatedR)
colnames(estDf) <- c("R")
estDf$day <- rownames(estDf)
lf <- loess(R ~ day, estDf)

plot(estDf$day, estDf$R, xlab="days", ylab="R", ylim=c(0,20), yaxt="n")
#lines(estDf$day, predict(lf), col = "blue")
abline(h=1, col="gray60")
axis(2, at=seq(0:max(estimatedR)))
```

The reproduction number has been reduced considerably over the last month or so, but is still above 1.
